---
title: "Homework 7"
author: "Joanna Yu (Fall 2018 Tuesday 4pm)"
date: "10/23/2018"
header-includes:
   - \usepackage{amsmath}
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The Meat

1a) Do you expect the distribution of this measure (ground beef consumption per capita per month) to be approximately normal? Why or why not?
$\\$
No, I do not expect the distribution of this to be normal because vegetarians consume zero meat. That will cause a spike at value 0. In a normal distribution, the tails should taper out. 
$\\$

1b) Do you expect the distribution of the sample mean to be approximately normal? Why or why not.
$\\$
Yes, I expect the distribution of the sample mean to be approximately normal. Since the sample size is sufficiently large and I do not expect the population to be very skewed, the distribution of the sample mean should be approximately normal. 
$\\$

1c) What is the 95% confidence interval for Berkeley students?
$$\begin{aligned}
(\bar x - 1.96*{\sigma\over\sqrt n} ,  \bar x + 1.96*{\sigma\over\sqrt n} ) \\ 
(2.45 - 1.96*{2\over \sqrt{100}}, 2.45 + 1.96*{2\over \sqrt{100}})\\
(2.058, 2.842)\\\\
\end{aligned}$$

GRE Scores

2. What is the real confidence level for the interval we have made, if the sample size is 10? What if the sample is 200?

If n=10, df=9
```{r}
1-2*pt(-1.96, df=9)
```
The real confidence level is 91.84% when the sample size is 10.

If n=200, df=199
```{r}
1-2*pt(-1.96, df=199)
```
The real confidence level is 94.86% when the sample size is 200.
$\\$

Maximum Likelyhood Estimation for an Exponential Distribution

3a) Write down the likelihood function, $L(\lambda)$.
$$
\begin{aligned}
  L(\lambda) &= f(x_1, \dots, x_n; \lambda) \\
       &= \prod_{i=1}^{n} f({x_i;\lambda}) \\
       &= \prod_{i=1}^{n} \lambda e^{-\lambda x_i} \\
\end{aligned}
$$

3b) Write down the log of the likelihood, and simplify it.
$$ \sum_{i=1}^n \left(log(\lambda)-\lambda x_i  \right) $$

3c) Take the derivative of the log of likelihood, set it equal to zero, and solve for $\lambda$. How is it related to the mean time between arrivals?
$$\begin{aligned}
\sum_{i=1}^n \frac{1}{\lambda}-x_i & = 0  \\
{1\over \lambda} &=\sum_{i=1}^n x_i\\
\lambda &= {1\over \sum_{i=1}^n x_i }
\end{aligned}$$

3d) Use R to plot the likelihood function. Then use optimize to approximate the maximum likelihood estimate for $\lambda$. How does your answer compare to your solution from part c?
```{r}
times = c(2.65871285, 8.34273228, 5.09845548, 7.15064545,
0.39974647, 0.77206050, 5.43415199, 0.36422211,
3.30789126, 0.07621921, 2.13375997, 0.06577856,
1.73557740, 0.16524304, 0.27652044)
lambda_likelihood <- function(lambda) {
 return(lambda^length(times) * exp(-lambda * sum(times)))
}
lambda <- seq(0, 1, by = 0.1)

library(ggplot2)
qplot(lambda,
      sapply(lambda, function(lambda) {lambda_likelihood(lambda)}),
      geom = 'line',
      main = 'Likelihood as a Function of lambda',
      xlab = 'lambda',
      ylab = 'Likelihood')
```

```{r}
optimize(lambda_likelihood, interval = lambda, maximum = TRUE)
1/mean(times)
```

Using optimize in R yields 0.3949072. The answer I have from c is 0.3949269.


