---
title: "Unit 9 Pre-Class Warm-Up"
author: "Joanna Yu (Tuesday 4pm Fall 2018)"
date: "10/29/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Q1. Is there a difference between the amount of money a senator raises and the amount spent?

Q1.1. Are the assumptions for a t-test met?

Yes, if we are looking at the difference bewteen the amount raised and spent, each random variable is identically distributed and independent. The distributions should be normal. 

$\\$
Q1.2. Is a paired test or an unpaired test more appropriate?

The paired test is more appropriate because we are trying to compare the difference for each senator. 
$\\$

Q1.3. 

```{r}
senator = read.csv("united_states_senate_2014_v2.csv")
summary(senator)
```

Q2. Do female Democratic senators raise more or less money than female Republican senators?

Q2.1. Are the assumptions for a t-test met?

Yes, the female Democratic senators and female Republican senators should be identically distributed and independent. The distributions should be normal. 

$\\$
Q2.2. Is a paired test or an unpaired test more appropriate?
An unpaired test is more appropriate because we are comparing the amount of money raised across two different groups of females.

$\\$

```{r}
```


Q3. Does the NRA prefer male senators or female senators?

Q2.1. Are the assumptions for a t-test met?

Yes, the NRA ratings for male and NRA ratings for female should be identically distributed and independent. The distributions should be normal.

$\\$
Q2.2. Is a paired test or an unpaired test more appropriate?
An unpaired test is more appropriate because we are comparing the NRA ratings between the two genders.

$\\$

```{r}
```

